\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{float}
\usepackage{enumitem}
\usepackage{caption}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    urlcolor=blue!60!black,
    citecolor=blue!60!black
}

\title{\textbf{MAEG3080 Fundamentals of Machine Intelligence}\\Final Project Report\\[0.5em]\large CIFAR-100 Image Classification with a Custom PyTorch Network}
\author{
    Student 1 Name (Student ID) \\
    Student 2 Name (Student ID) \\
    Student 3 Name (Student ID)
}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
This report presents our final project for MAEG3080: training an image classifier from scratch on CIFAR-100 using PyTorch, without pretrained weights. We use a compact SE-ResNet-style architecture with squeeze-and-excitation blocks. We first train a short baseline with basic augmentations, then finetune from those weights with RandAugment and a held-out validation split. The final model achieves \textbf{65.17\%} accuracy on the official test set, evaluated with an independent \texttt{results.py} script as required by the project.
\end{abstract}

\section{Introduction}
The objective is to train a high-performing classifier on CIFAR-100 ($100$ classes, $32 \times 32$ RGB images). With no pretrained weights, performance depends on architecture choice, augmentation, and optimization. We prioritize a \emph{reproducible, efficient} pipeline (Google Colab + Google Drive checkpoints) and report official test accuracy from a dedicated evaluation script.

\section{Dataset}
CIFAR-100 contains $60{,}000$ images: $50{,}000$ for training and $10{,}000$ for testing. We normalize inputs with the standard channel statistics:
\[
\mu = (0.5071, 0.4867, 0.4408), \quad
\sigma = (0.2675, 0.2565, 0.2761).
\]
For finetuning, we hold out $10\%$ of the training labels as a stratified validation set (seed $42$); the baseline stage uses the full $50{,}000$ training images and evaluates on the test set each epoch for comparability with a historic short baseline.

\section{Methodology}
\subsection{Model Architecture}
We use a \textbf{small SE-ResNet} (\texttt{se\_resnet} in our codebase): residual stages with \textbf{squeeze-and-excitation} channel attention, \texttt{base\_width}${}=32$, \textbf{two} residual blocks per stage (\texttt{2$\times$2$\times$2$\times$2}), and global average pooling plus a linear classifier ($100$ classes). This design balances capacity and training time on commodity GPUs.

\subsection{Training Strategy}
We use a \textbf{two-stage} procedure:
\begin{itemize}[leftmargin=1.2em]
    \item \textbf{Stage A (baseline):} $4$ epochs on full training data, \textbf{basic} augmentation (crop, flip), \textbf{no} Mixup/CutMix, SGD (lr $0.1$), warmup $1$ epoch, cosine decay, test-set evaluation each epoch. Produces a $\sim$33\% test checkpoint used as initialization.
    \item \textbf{Stage B (finetune):} Load weights only (\texttt{--init-from}), \textbf{new} optimizer and cosine schedule. \textbf{RandAugment} ($N{=}2$, $M{=}9$), random erasing ($p{=}0.05$), label smoothing $0.05$, SGD with initial lr $0.05$, warmup $5$ epochs, $30$ epochs total. \textbf{EMA disabled} in both stages for stability. Evaluation during finetuning uses the \textbf{validation} split; the \textbf{official test} is reserved for \texttt{results.py}.
\end{itemize}
Training uses automatic mixed precision on CUDA where applicable (\texttt{torch.autocast}); checkpoints are written to disk under a configurable save directory.

\subsection{Implementation Details}
Implementation is in PyTorch (\texttt{train.py}, \texttt{model.py}, \texttt{utils.py}). The Colab path runs \texttt{scripts/quick\_se\_resnet\_results.sh}: baseline if needed, finetune, then \texttt{results.py} on the best finetune checkpoint for the \textbf{final test number}.

\section{Experiments}
\subsection{Experimental Setup}
Table~\ref{tab:hyperparameters} summarizes the two stages. Table~\ref{tab:results} reports accuracy.

\begin{table}[H]
    \centering
    \caption{Hyperparameters for the two-stage SE-ResNet pipeline (final submission).}
    \label{tab:hyperparameters}
    \begin{tabular}{lll}
        \toprule
        Setting & Stage A (baseline) & Stage B (finetune) \\
        \midrule
        Epochs & 4 & 30 \\
        Batch size & 64 & 256 \\
        Augmentation & Basic (crop, flip) & RandAugment + erasing \\
        Mixup / CutMix & None & None \\
        Optimizer & SGD (momentum $0.9$, WD $5\times 10^{-4}$) & Same \\
        LR schedule & Warmup 1 + cosine & Warmup 5 + cosine \\
        Initial LR & $0.1$ & $0.05$ \\
        Label smoothing & $0$ & $0.05$ \\
        Eval during training & Test ($10{,}000$) & Val ($10\%$ of train) \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Results}
\begin{table}[H]
    \centering
    \caption{CIFAR-100 accuracy. Official test is from \texttt{results.py} on the finetune checkpoint.}
    \label{tab:results}
    \begin{tabular}{lll}
        \toprule
        Stage & Metric & Value \\
        \midrule
        A: baseline ($4$ epochs) & Test accuracy & $\approx 33.3\%$ \\
        B: finetune ($30$ epochs) & Best validation & $65.22\%$ \\
        B: finetune (final) & \textbf{Official test} & \textbf{65.17\%} \\
        \bottomrule
    \end{tabular}
\end{table}

Validation and test accuracies are aligned ($\sim$65.2\%), indicating that the stratified val split is a reasonable proxy and we do not observe extreme overfitting to the validation fold.

\subsection{Training Curve}
Loss decreases and accuracy increases monotonically over the short baseline; finetune metrics improve steadily over $30$ epochs. Plots can be generated from \texttt{history.json} in each run directory if required.

\section{Discussion}
\begin{itemize}[leftmargin=1.2em]
    \item \textbf{Initialization matters:} Starting from the Stage~A weights gave a large jump over training the finetune recipe from scratch in the same wall-clock budget.
    \item \textbf{RandAugment + longer schedule:} Stronger augmentation and more epochs on top of a sane baseline dominated our earlier attempts with a much larger WRN trained from random initialization in limited time.
    \item \textbf{Trade-offs:} The SE-ResNet path is faster to iterate than a full Wide-ResNet Hydra ladder; the cost is a lower ceiling than very large models trained for hundreds of epochs.
\end{itemize}

\section{Group Contribution}
Clearly explain how each member contributed to the project.

\begin{itemize}[leftmargin=1.2em]
    \item Student 1: Data pipeline, augmentation experiments, report writing
    \item Student 2: Model design, training implementation, checkpoint management
    \item Student 3: Hyperparameter tuning, presentation slides, result analysis
\end{itemize}

\section{Conclusion}
We trained a compact SE-ResNet from scratch on CIFAR-100 using a two-stage baseline-plus-finetune workflow in PyTorch, with RandAugment and careful scheduling on the second stage. Our best model achieves \textbf{65.17\%} test accuracy via \texttt{results.py}. Future work could explore wider networks or longer schedules if compute allows.

\section*{References}
\begin{enumerate}[leftmargin=1.5em]
    \item A. Krizhevsky, ``Learning Multiple Layers of Features from Tiny Images,'' 2009.
    \item K. He, X. Zhang, S. Ren, and J. Sun, ``Deep Residual Learning for Image Recognition,'' 2016.
    \item J. Hu, L. Shen, and G. Sun, ``Squeeze-and-Excitation Networks,'' 2018.
    \item TorchVision and PyTorch official documentation.
\end{enumerate}

\end{document}
